Keep your Laziness in Check

### 2018-02-20 paper/main.tex
[paper] backup, finished spec for rotate but the text around it still needs much work

\leo{The singular feels weird here. Is it just me?}
As a first example, consider purely functional queues, a very
popular data structure in Haskell. Programmers typically implement
such queues with two lists---a front list holding the queue elements
in FIFO order, and a back list holding the other elments in reversed
order. When the front list is empty, a dequeue operation requires a
reversal of the back list to refill the front list. These specific
dequeue operations take $O(n)$ time to finish, and are much slower
than their imperative counterparts. This occasional performance
degradation is compounded by persistence (where the $O(n)$ operation
could be repeated many times), and can be unacceptable in real-time
systems.

Since \lk{rotate} is functionally equivalent to \lk{front ++ (reverse
back)}, a programmer could have implemented it naively and never
discover the mistake through traditional testing! The two only differ
in their laziness behavior, and there is no way of specifying such
subtle differences in any testing frameworks today.

@@

In this section, we will discuss how to specify the laziness behaviors
of Haskell functions through running examples of purely functional
maps from integers to integers. The map operations of interest
are \lk{insert} and \lk{lookup} with the following types:
\begin{inlinecode}
insert :: IntMap -> Int -> Int -> IntMap Int
lookup :: IntMap -> Int        -> Maybe  Int
\end{inlinecode}
%
The \lk{IntMap} can be implemented with a few different laziness
flavors. We can choose a binary search tree representation for
efficient lookups, or we can choose an association list representation
for easy inserts. The binary search tree representation necessarily
evaluates the keys of the map to order the keys, but it can be lazy in
the values. On the other hand, the association list representation can
be lazy in the keys as well. To summarise, we have maps that are
strict in both keys and values, maps that are strict in keys but lazy
in values, and maps that are fully lazy in both keys and values.

In Haskell, triggering the evaluation of a lazy function requires
exerting some demand on its result. Such demands can be pattern
matches, or further computations that depend on the result. In typical
Haskell programming, such demands are exerted by the contextual code
surrounding the result of a lazy function. We will use the
phrase \textit{demand context} to refer to functions that causes
evaluation of lazy values.

An evaluation of a lazy data structure can partially explore the lazy
structure's shape. For example, \lk{lookup} on an association list
based map evaluates the association list until it finds the first
matching key. If such a key is halfway inside the association list,
then the values past this key will not be evaluated in this operation.
%
To describe how much of a data structure is evaluated, we will
use \textit{demand types}. A \textit{demand type} is almost isomorphic
to the corresponding data type, except we will wrap all the
constructor arguments in a \lk{Thunk} to represent the possible lack
of evaluation. For example, consider the \lk{Tree} type and
its \textit{demand type} \lk{TreeD}.

>>>>

@@

As a first example, consider purely functional queues, a very
popular data structure in Haskell. Programmers typically implement
such queues with two lists---a front list holding the queue elements
in FIFO order, and a back list holding the other elments in reversed
order. When the front list is empty, a dequeue operation requires a
reversal of the back list to refill the front list. These specific
dequeue operations take $O(n)$ time to finish, and are much slower
than their imperative counterparts. This occasional performance
degradation is compounded by persistence where the $O(n)$ operation
could be repeated many times, and can be unacceptable in real-time
systems.

Since \lk{rotate} is functionally equivalent to \lk{front ++ (reverse
back)}, a programmer could have implemented it naively and never
discover the mistake through traditional testing! The two only differ
in their laziness behavior. Fortunately, we can use StrictCheck to
write a specification that makes this property of \lk{rotate}
testable.

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
%
\begin{figure}[H]
\centering
\begin{inlinecode}
data ListD a = NilD
             | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
\caption{\lk{ListD} type}
\label{fig:ListD}
\end{figure}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and a nullary constructor \lk{T}
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor
from \lk{Thunk}. For example, given the list \lk{[1, 2, 3]}, the
value \lk{ConsD T (ConsD T T)} represents the \textit{demand} that
reveals the first 2 \lk{Cons} constructors of the list, but does not
evaluate any other parts of the list. To improve readability of demand
values, we will also use the syntax \lk{[\_, \_, ...]} where
appropriate.

In Haskell, to trigger the evaluation of a lazy function, we need to
exert some \textit{demand} on the result of that function. This
suggests a natural flow for the specification of a lazy function ---
given the demand on the output, the specification should calculate the
demands on the input. However, this is not enough in general because
the demand on the input could depend on the value of the inputs as
well. The \lk{take} function introduced in \Cref{sec:intro} is such a
case since the length of the output from \lk{take} depends on its
integer argument.

Putting these pieces together, we arrive at an appropriate type for
the specification of \lk{rotate}: \lk{ListD a -> [a] -> [a] -> [a] ->
(ListD a, ListD a, ListD a)}.
%
\hzh{The description here needs improvement}
To specify the laziness behavior of \lk{rotate}, we need to inspect
the 4 different cases of \lk{rotate}. When both \lk{fs} and \lk{bs}
are empty, the accumulator is returned directly; when \lk{fs} is an
empty list, \lk{rotate} immediately recurses on the tail of \lk{bs};
when \lk{bs} is empty, \lk{rotate} produces a new list with the
original head of \lk{fs} as the new head, and the tail of that list is
the result of a recursion; finally, when both \lk{fs} and \lk{bs} are
not empty, \lk{rotate} also immediately produces a new list with the
appropriate head, and conses \hzh{is this a word} the original head
of \lk{bs} onto the accumulator.

Turning our intuition above into a Haskell function, we complete the
specification of \lk{rotate} in \Cref{fig:rotate-spec}
\begin{figure}[H]
\centering
\begin{inlinecode}
rotate_spec :: ListD a
            -> [a] -> [a] -> [a]
            -> (Thunk (ListD a), Thunk (ListD a), Thunk (ListD a))
rotate_spec NilD                     fs bs as = (E NilD, E NilD, E NilD)
rotate_spec dOut@(ConsD dElem dTail) fs bs as =
  case (fs, bs) of
    ([],           []) ->
      (E NilD, E NilD, E dOut)
    ([],      b : bs') ->
      let (_, dBs', dBAs) = rotate_spec dTail [] bs' (b:as)
      in case dBAs of
           T                -> (E NilD, T, T)
           E (ConsD dB dAs) -> (E NilD, ConsD dB dBs', dAs)
           E NilD           -> error ``Impossible''
    (_ : fs',      []) ->
      let (dFs', _, dAs) = rotate_spec dTail fs' [] as
      in  (E (ConsD dElem dFs'), E NilD, dAs)
    (_ : fs', b : bs') ->
      let (dFs', dBs', dBAs) = rotate_spec dTail as' bs' (b:as)
          dFs                = E (ConsD dElem dFs')
      in case dBAs of
           T                -> (dFs, T, T)
           E (ConsD dB dAs) -> (dFs, E (ConsD dB dBs'), dAs)
           E NilD           -> error ``Impossible''
\end{inlinecode}
\caption{Specification for \lk{rotate}}
\label{fig:rotate-spec}
\end{figure}
